---
title: "R Notebook"
output: html_notebook
---



```{r}
#install.packages("faraway")
#install.packages("tscount")
#install.packages("pscl")


#library(faraway); library(MASS); library(pscl)

#data
data("discoveries")
```


Problem 1:  The discoveries dataset in the faraway package lists the numbers of "great" inventions and scientific discoveries in each year from 1860-1959.  Investigate the number of discoveries over time.  Use the appropriate Poisson model to fit the number of discoveries against the year.  Hint you may need to create this.  Comment on any findings of the model you have.
```{r}
head(discoveries)

summary(discoveries)
plot(discoveries, main = "Number of Discoveries")
```
Some exploration of the data
```{r}
#test whether discoveries is a time series object
ts(discoveries) 
time(discoveries)

```
Frequency = 1 => yearly data 


A plot of number of discoveries vs time shows some increase from 1860 to early 1900's with a peak in number of discoveries in the early 1880's. The average number of discoveries remains around the total mean of 3.1 per year until about 1930 when we see a steady decline until 1959. 
```{r}
mean(discoveries)
plot(discoveries, main = "Number of Discoveries")
abline(mean(discoveries), 0, col = "red")

```
We calculate the mean of number of discoveries pre and post 1930 and see a clear step down from 3.6 to 1.9 per year. 

```{r}
#check mean pre and post 1930
disc_pre1930 <- window(discoveries, start = 1860, end = 1930)
disc_post1930 <- window(discoveries, start = 1930, end = 1959)

mean(disc_pre1930)
mean(disc_post1930)

```

```{r}
#plots with both means
par(mfrow = c(1, 2))
plot(disc_pre1930, main = "Discoveries pre-1930", ylim = c(0, 12))
abline(mean(disc_pre1930), 0, col = "red")

plot(disc_post1930, main = "Discoveries post-1930", ylim = c(0, 12))
abline(mean(disc_post1930), 0, col = "blue")

```

We first fit a linear regression model. The negative slope indicates # of discoveries reduces with time. The $r^2$ is very small indicating the poor fit of the model as expected. We see some curvature in the residual plots.
```{r}
n_discoveries <- as.numeric(discoveries)
d_discoveries = data.frame(n_discoveries)
d_discoveries$year = 1860:1959

model0 <- lm(n_discoveries ~ year, data = d_discoveries)
summary(model0)

```


```{r}
par(mfrow = c(2, 2))
plot(model0)

```

We plot the predicted values. The linear model is consistent with the initial observation that number of discoveries is decreasing with time. 
```{r}
#calculate model predictions
model0_pred <- predict(model0)

#transform into time series
disc_pred <- ts(model0_pred, start = c(1860, 1), end = c(1959, 1), frequency = 1)

plot(discoveries)
points(disc_pred, pch = 3, col = "red")

```



We fit a poisson regression model since this is a count type data. The model is is $log(μi hat) = 0.25696 + 0.22869 xi$. The coefficient for x is significant given its small p-value; since it is > 0, the greater the time, the larger the number of expected discoveries, or for one unit increase of time (1 year), the number of discoveries will increase and get multiplied by 1.26. The coefficient for x is the multiplicative term to calculate the estimated number of discoveries when time increases by 1 unit (1 year). See a significant reduction in deviance from the null model. 
It seems that the coefficient should be < 0 to indicate a reduction in discoveries with time. We also see that the residual deviance is much greater than the degrees of freedom. 
```{r}
model1 <- glm(discoveries ~ ., family = poisson(link = "log"), data = discoveries)
summary(model1)

```

Is the model a good fit? The p-value is large indicating no evidence of a lack of fit. We can use ANOVA to compare the current model with a null model and we observe the same conclusion. 

```{r}
1 - pchisq(deviance(model1), df.residual(model1))

anova(model1, test = "Chi")
```


Model diagnostics. We see the same outlier point 26. We also see some non-constant variance in the residuals
```{r}
par(mfrow = c(2, 2))
plot(model1)
```
We can check if we have an issue with dispersion by plotting the relationship between the mean and variance. We see some dispersion of the data and this may be the reason for the poor fit since for a poisson model to be valid the mean and the variance should be the same.  

```{r}
model1_mean <- log(fitted(model1))
model1_var <- log((discoveries-fitted(model1))^2)

par(mfrow = c(1, 2))
plot(model1_mean, model1_var)
abline(0,1, col = "blue")
```

How well is the model predicting? Visually the model appears to do a good job at prediction. We see that the model predicts slightly higher than the lowest actuals and slightly lower than the highest actuals. The mean of the predicted values is the same as the mean of the actual number of discoveries
```{r}
#calculate model predictions
model1_pred <- predict(model1, type = "response")

#transform into time series
disc_pred1 <- ts(model1_pred, start = c(1860, 1), end = c(1959, 1), frequency = 1)

plot(discoveries)
points(disc_pred1, pch = 20, col = "red")

plot(d_discoveries$n_discoveries, model1_pred)
```

```{r}
#compare mean of prediction with mean of the data
mean(disc_pred1)
mean(discoveries)
```

Since there are some years with no discoveries we'll use the zero inflated count model to attempt a better fit. 
```{r}
#numeric data 
d_discoveries
d_discoveries$index <- seq(1, 100, length.out = 100)

#fit the model 
model_zinf <- zeroinfl(n_discoveries ~ index, d_discoveries)
summary(model_zinf)

#calculate model predictions
model_zinf_pred <- predict(model_zinf, type = "response")

#transform into time series
disc_zinf <- ts(model_zinf_pred, start = c(1860, 1), end = c(1959, 1), frequency = 1)

#plot prediction 
plot(discoveries)
points(disc_zinf, pch = 20, col = "red")

#plot residuals
plot(residuals(model_zinf))
abline(0, 0, col = "blue")
```
```{r}
#lastly we try a hurdle model 
model_h <- hurdle(n_discoveries ~ index, d_discoveries)
summary(model_h)

#calculate model predictions
model_h_pred <- predict(model_h, type = "response")

#transform into time series
disc_h <- ts(model_h_pred, start = c(1860, 1), end = c(1959, 1), frequency = 1)

#plot prediction 
plot(discoveries)
points(disc_h, pch = 20, col = "red")

#plot residuals
plot(residuals(model_h))
abline(0, 0, col = "blue")
```










